Cover Your Basis: Comprehensive Data-Driven Characterization of the Binary Black Hole Population

### 2022-12-05 src/tex/intro.tex
update with salvos comments

Observations of gravitational waves (GWs) from compact binary mergers are becoming a regular occurrence, 
producing a catalog of events that recently surpassed 90 such detections \citep{GWTC1,gwtc2,GWTC3}. As the catalog continues to grow, so does our understanding of the underlying astrophysical population of compact binaries \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
Following numerous improvements to the detectors since the last observing run, the anticipated sensitivities for the upcoming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration suggest detection rates as high as once per \emph{day} \citep{aLIGO,aVIRGO,LVK_prospects,KAGRAProg}. With the formation history of these dense objects encoded in the details of their distribution \citep{Rodriguez_2016,Farr2017Nature,Zevin_2017,Farr_BinnedSpin}, the likely doubling in size of the catalog with the next observing run could provide another leap in our understanding of compact binary astrophysics. 
Beyond formation physics, population-level inference of the compact binary catalog has also been shown to provide novel measurements of cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo,JoseSpectralSirens}, constrain modified gravitational wave propagation \citep{OkounkovaBirefringence,ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, search for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and to constrain physics of neutron stars \citep{Golomb_EOS,LandryRead_NS_Masses2021}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

>>>>

Observations of gravitational waves (GWs) from compact binary mergers are becoming a regular occurrence, 
producing a catalog of events that recently surpassed 90 such detections \citep{GWTC1,gwtc2,GWTC3}. As the catalog continues to grow, so does our understanding of the underlying astrophysical population of compact binaries \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
Following numerous improvements to the detectors since the last observing run, the anticipated sensitivities for the upcoming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration suggest detection rates as high as once per \emph{day} \citep{aLIGO,aVIRGO,LVK_prospects,KAGRAProg}. With the formation history of these dense objects encoded in the details of their distribution \citep{1503.04307,Rodriguez_2016,Farr2017Nature,Zevin_2017,Farr_BinnedSpin}, the likely doubling in size of the catalog with the next observing run could provide another leap in our understanding of compact binary astrophysics. 
Beyond formation physics, population-level inference of the compact binary catalog has also been shown to provide novel measurements of cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo,JoseSpectralSirens}, constrain modified gravitational wave propagation \citep{OkounkovaBirefringence,ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, search for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and to constrain physics of neutron stars \citep{Golomb_EOS,LandryRead_NS_Masses2021}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

### 2022-12-05 src/tex/results.tex
update with salvos comments

The marginal mass ratio distribution inferred by the B-spline model is shown in figure \ref{fig:q_distribution}.  These results suggest we may be seeing the first signs of departure from a simple power law behavior.  We find a potential plateau or decrease in the merger rate near equal mass ratios, as well as a broader tail towards unequal mass ratios 
than the power law based models find. Our results also suggest a shallower slope from $q\sim0.3$ to $q\sim0.7$, though uncertainty is larger in this region. The sharp decrease in rate just below 
$q\sim0.5$ is due to the minimum mass ratio truncation defined by $q_\mathrm{min}=\frac{m_\mathrm{min}}{m_1}$. When marginalizing over the primary mass 
distribution with a strong peak at $10\msun$, the mass ratio distribution truncates at $q\sim0.5$: the minimum mass, $5\msun$, 
divided by the most common primary mass, ${\sim}10\msun$.

@@

We model the spin magnitude distributions as IID B-Spline distributions. 
Figure \ref{fig:iid_spinmag_dist} shows the inferred spin magnitude distribution with the B-Spline model, compared with the \textsc{Default} model from \citet{o3b_astro_dist}. 
The B-Spline model results are consistent with those using the Beta distribution, peaking near $a\sim0.2$, with 90\% of BBH spins below 
\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][a_90percentile]}$} at 90\% credibility. The B-Spline model does not impose vanishing support at the extremal values like the Beta distribution, 
allowing it to probe the zero-spin question. We find broad support, with large variance, for non-zero probabilities at $a_i=0$, but cannot confidently determine the presence of 
a significant non-spinning subpopulation, corroborating similar recent conclusions \citep{BuildBetterSpinModels,Callister_NoEvidence,GWTC3MonashSpin,Mould2022}. 
We repeat the same analysis with independent B-Spline distributions for each spin magnitude component. In figure \ref{fig:ind_spinmag_dist} 
we show the inferred primary (orange), and secondary (olive) spin magnitude distributions inferred when relaxing the IID assumption. We find that the primary spin magnitude distribution 
peaks slightly higher, at $a\sim0.25$, than the IID B-Spline model, but with similar support at near vanishing spins. The secondary spin magnitude distribution is more uncertain due to the 
higher measurement uncertainty when inferring the secondary spins of BBH systems. The secondary distribution also peaks at smaller spin magnitudes of $a\sim0.15$, showing potentially rates at $a\sim0$ than the primary distribution or B-Spline IID spin magnitude distribution in figure \ref{fig:iid_spinmag_dist}, though uncertainties are large. While the distributions are broadly consistent, we could be seeing signs that component spin magnitude distributions are uniquely distributed, which can be produced through mass-ratio reversal in isolated binary evolution \citep{Mould2022}.

@@

Recent analysis of the GWTC-3 BBH population has shown evidence for an increasing merger rate with redshift, nearly ruling out a merger rate that is 
constant with co-moving volume \citep{Fishbach_2018redshift,o3b_astro_dist}. When extending the power law form of the previously used model to have a modulation 
that we model with B-Splines, the merger rate as a function of redshift in figure \ref{fig:rofz} shows mild support for features departing from the underlying power law. 
In particular, we see a small increase in merger rate from $z\sim0.09$ to $z\sim0.2$ (where we best constrain the rate), followed by a plateau in the rate from $z\sim0.2$ to $z\sim0.4$. 
At larger redshifts, where we begin to have sparse observations, we see no sign of departure from the power-law as the rate continues to increase with redshift. 
The underlying power-law slope of our B-spline modulated model is consistent with the GWTC-3 results with the underlying model by itself: the \textsc{PowerlawRedshift} model 
found $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} when inferred with the \textsc{PowerlawPeak} mass, and \textsc{Default} spin models. Our more 
flexible model infers a power law slope of $\lambda_z = $\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][lamb]}$}. We show the basis spline modulations or departure 
from the power law in \ref{fig:z_modulation} compared to the prior -- showing where we cannot constrain any significant deviations from the simpler parametric power law model. 
The extra freedom of our model does inflate the uncertainty in its rate estimates, especially at $z\sim0$ where there are not any observations in the catalog. 
We find a local ($z=0$) merger rate of $\mathcal{R}_0 = $\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][local_rate]}\,Gpc^{-3}yr^{-1}$} using the B-Spline modulation model 
which compares to $\mathcal{R}_0 = $\result{$\CIPlusMinus{\macros[PLPeak][local_rate]}\,Gpc^{-3}yr^{-1}$} for the GWTC-3 result.

>>>>

The marginal mass ratio distribution inferred by the B-spline model is shown in figure \ref{fig:q_distribution}.  These results suggest we may be seeing the first signs of departure from a simple power law behavior.  We find a potential signs of a plateau or decrease in the merger rate near equal mass ratios, as well as a broader tail towards unequal mass ratios 
than the power law based models find, although a smooth power law is still consistent with these results given the large uncertainties. Our results also suggest a shallower slope from $q\sim0.3$ to $q\sim0.7$, though uncertainty is larger in this region. The sharp decrease in rate just below 
$q\sim0.5$ is due to the minimum mass ratio truncation defined by $q_\mathrm{min}=\frac{m_\mathrm{min}}{m_1}$. When marginalizing over the primary mass 
distribution with a strong peak at $10\msun$, the mass ratio distribution truncates at $q\sim0.5$: the minimum mass, $5\msun$, 
divided by the most common primary mass, ${\sim}10\msun$.

@@

We model the spin magnitude distributions as IID B-Spline distributions. 
Figure \ref{fig:iid_spinmag_dist} shows the inferred spin magnitude distribution with the B-Spline model, compared with the \textsc{Default} model from \citet{o3b_astro_dist}. 
The B-Spline model results are consistent with those using the Beta distribution, peaking near $a\sim0.2$, with 90\% of BBH spins below 
\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][a_90percentile]}$} at 90\% credibility. The B-Spline model does not impose vanishing support at the extremal values like the Beta distribution, 
allowing it to probe the zero-spin question. We find broad support, with large variance, for non-zero probabilities at $a_i=0$, but cannot confidently determine the presence of 
a significant non-spinning subpopulation, corroborating similar recent conclusions \citep{BuildBetterSpinModels,Callister_NoEvidence,GWTC3MonashSpin,Mould2022}. 
We repeat the same analysis with independent B-Spline distributions for each spin magnitude component. In figure \ref{fig:ind_spinmag_dist} 
we show the inferred primary (orange), and secondary (olive) spin magnitude distributions inferred when relaxing the IID assumption. We find no signs that the spin magnitude distributions are not IID but that the primary spin magnitude distribution peaks slightly higher, at $a\sim0.25$, than the IID B-Spline model in figure \ref{fig:iid_spinmag_dist}, but with similar support at near vanishing spins. The secondary spin magnitude distribution is more uncertain due to the 
higher measurement uncertainty when inferring the secondary spins of BBH systems \citep{1403.0129,1611.01122}. The secondary distribution also peaks at smaller spin magnitudes of $a\sim0.15$, showing potentially rates at $a\sim0$ than the primary distribution or B-Spline IID spin magnitude distribution in figure \ref{fig:iid_spinmag_dist}, though uncertainties are large. While the distributions are broadly consistent, we could be seeing signs that component spin magnitude distributions are uniquely distributed, which can be produced through mass-ratio reversal in isolated binary evolution \citep{Mould2022}.

@@

Recent analysis of the GWTC-3 BBH population has shown evidence for an increasing merger rate with redshift, nearly ruling out a merger rate that is 
constant with co-moving volume \citep{Fishbach_2018redshift,o3b_astro_dist}. When extending the power law form of the previously used model to have a modulation 
that we model with B-Splines, the merger rate as a function of redshift in figure \ref{fig:rofz} shows mild support for features departing from the underlying power law. 
In particular, we see a small increase in merger rate from $z\sim0.09$ to $z\sim0.2$ (where we best constrain the rate), followed by a plateau in the rate from $z\sim0.2$ to $z\sim0.4$. 
At larger redshifts, where we begin to have sparse observations, we see no sign of departure from the power-law as the rate continues to increase with redshift. 
The underlying power-law slope of our B-spline modulated model is consistent with the GWTC-3 results with the underlying model by itself: the \textsc{PowerlawRedshift} model 
found $\lambda_z = $\result{$\CIPlusMinus{\macros[PLPeak][lamb]}$} when inferred with the \textsc{PowerlawPeak} mass, and \textsc{Default} spin models. Our more 
flexible model infers a power law slope of $\lambda_z = $\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][lamb]}$}. We show the basis spline modulations or departure 
from the power law in \ref{fig:z_modulation} compared to the prior -- showing where we cannot constrain any significant deviations from the simpler parametric power law model. 
The extra freedom of our model does inflate the uncertainty in its rate estimates, especially at $z\sim0$ where there are not any observations in the catalog. 
We find a local ($z=0$) merger rate of $\mathcal{R}_0 = $\result{$\CIPlusMinus{\macros[BSplineIIDCompSpins][local_rate]}\,\mathrm{Gpc}^{-3}\mathrm{yr}^{-1}$} using the B-Spline modulation model 
which compares to $\mathcal{R}_0 = $\result{$\CIPlusMinus{\macros[PLPeak][local_rate]}\,\mathrm{Gpc}^{-3}\mathrm{yr}^{-1}$} for the GWTC-3 result.
